---
title: "Anthro analysis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
library(lme4)
library(tidyverse)
library(magrittr)
library(gridExtra)
library(binom) # For AC binomial confidence intervals
```

```{r, echo=F, include=F, cache=T}
load("~/Library/CloudStorage/OneDrive-HarvardUniversity/School/Harvard/HaSET/Code/anthro_data.RData")

##### Exploratory analysis and data wrangling #####
dat <- anthro %>% subset(age >= 0 & age < 1000 & !is.na(length)) # Remove invalid age and missing lengths - 13418 observations

# Categorizing age 
dat %<>% mutate(visit = case_when(age <= 3 ~ "Birth", 
                                  age > 3 & age < 10 ~ "Day 6",
                                  age > 20 & age < 35 ~ "Day 28",
                                  age > 35 & age < 50 ~ "Day 42",
                                  age > 165 & age < 200 ~ "6 months",
                                  age > 350 & age < 380 ~ "12 months",
                                  age > 715  ~ "24 months")) %>% 
  mutate(visit = factor(visit,levels = c("Birth","Day 6","Day 28","Day 42","6 months","12 months","24 months")))

# Convert age to months
dat %<>% mutate(age=age/30.437)
# Create standardized age variable
dat %<>% mutate(age_s = scale(age))
# Remove unnecessary variables
dat %<>% select(-c(cdob, intdt, dcid)) 
# Number of observations per child + relabel ID
dat %<>% group_by(cuuid) %>% mutate(n_obs = n(), cuuid=factor(cur_group_id())) %>% ungroup()
# Subset WHO standard data
who_sd <- who %>% filter(who=="-2 SD")

# Add stunted column
isStunted <- Vectorize(function(age_in, length_in){
  who_standard <- who_sd[which.min(abs(who_sd$age-age_in)) ,"standard"] # Get corresponding standard
  return(length_in < who_standard)
})
dat %<>% mutate(stunt=isStunted(age,length))

# Remove facility visits after birth - 13418 to 11086 observations
dat %<>% filter(location=="community" | visit=="Birth")

# Append splines to dataset
dat %<>% mutate(spline1 = case_when(age > 0 & age <= 1 ~ age, age==0 ~ 0, TRUE ~ 1),
                spline2 = case_when(age > 1 & age <= 6 ~ age-1, age <= 1 ~ 0, TRUE ~ 5),
                spline3 = case_when(age > 6 & age <= 12 ~ age-6, age <= 6 ~ 0, TRUE ~ 6),
                spline4 = case_when(age > 12 & age <= 29 ~ age-12, age <= 12 ~ 0, TRUE ~ 12))




##### Modeling #####
# LME with random intercept
mod1 <- lmer(length ~ age_s + (1|cuuid), data=dat, REML=F) 

# LME with random intercept+slope
mod2 <- lmer(length ~ age_s + (1+age_s|cuuid), data=dat, REML=F) 

# LME with random intercept and quadratic age
mod3 <- lmer(length ~ age_s + I(age_s^2) + (1|cuuid), data=dat, REML=F)

# LME with random intercept+slope and quadratic age
mod4 <- lmer(length ~ age_s + I(age_s^2) + (1 + age_s|cuuid), data=dat, REML=F)

# LME with random intercept and piecewise-linear age
mod5 <- lmer(length ~ spline1 + spline2 + spline3 + spline4 + (1|cuuid), data=dat, REML=F)

# LME with random intercept+slope and piecewise-linear age
mod6 <- lmer(length ~ spline1 + spline2 + spline3 + spline4 + (1 + age_s|cuuid), data=dat, REML=F)





##### Removing unreliable points #####
# Function to remove non-monotonically increasing points
removeMI <- function(dat.in){
  id.rm <- which(dat.in$length - lag(dat.in$length) < 0) # Function to remove non-increasing rows
  
  while (length(id.rm) > 0){
    dat.in <- slice(dat.in,-id.rm)
    id.rm <- which(dat.in$length - lag(dat.in$length) < 0) 
  }
  return(dat.in)
}
dat.mi <- dat %>% group_by(cuuid) %>% group_modify(~ removeMI(.)) # n=11657 after removal

# Function that removes points not within 1.5sd, given the model
removesd <- function(mod.in, dat.in){
  sd_val <- summary(mod.in)$sigma
  dat.sd <- dat.in %>% select(cuuid,age,age_s,length,visit,n_obs) %>% 
    mutate(p_length = predict(mod.in)) %>%
    filter(length >= (p_length-1.5*sd_val) & length <= (p_length+1.5*sd_val))
  return(dat.sd)
}

# Function that averages out values if more than one observation per individual in a given time window
averageMulti <- function(dat.in){
  # If more than one observation in a window, average them out
  if (nrow(dat.in) > 1){
    avg <- dat.in %>% select(age,length) %>% apply(2,mean)
    dat.out <- data.frame(t(avg))
    return(dat.out)
  }else{
    return(dat.in)
  }
}

# Function that calculates prevalence for the given data
prevCalc <- function(dat.in, modelname=""){
  dat.in <- dat.in %>% group_by(cuuid,visit) %>% group_modify(~ averageMulti(.)) %>% ungroup()
  pred.stunt <- dat.in %>% select(age,length,visit) %>% mutate(stunt=isStunted(age,length))
  prop <- table(pred.stunt$visit, pred.stunt$stunt) %>% prop.table(1) %>% .[,2]
  out <- tibble(c("Birth","Day 6","Day 28","Day 42","6 months","12 months","24 months"), prop, modelname)
  colnames(out) <- c("Category", "Proportion", "Model")
  return(out)
}

# Function to calculate incidence and reversal
# Baseline determines if we are calculating incidence/reversal based on shifting time points (6 days vs. 6 months, 6 months vs 12 months) or a baseline time point (6 days vs 6 months, 6 days vs 12 months)
# To select a baseline, pass in one of the categories: "Birth","Day 6","Day 28","Day 42","6 months","12 months","24 months"
inc_and_rev_Calc <- function(dat.in, modelname="", baseline=NULL){
  # Average out multiple instances per window
  dat.in <- dat.in %>% group_by(cuuid,visit) %>% group_modify(~ averageMulti(.)) %>% ungroup()
  pred.stunt <- dat.in %>% select(cuuid,age,length,visit) %>% mutate(stunt=isStunted(age,length))
  
  # Setup storage output
  selectID.inc <- selectID.rev <- pred.stunt %>% pull(cuuid)
  # Choose the categories we are interested in
  categories <- c("Day 6","6 months","12 months","24 months")
  #categories <- c("Birth","Day 6","Day 28","Day 42","6 months","12 months","24 months")
  
  out.inc <- data.frame(matrix(ncol = length(categories), nrow = 0))
  out.rev <- data.frame(matrix(ncol = length(categories), nrow = 0))
  colnames(out.inc) <- categories
  colnames(out.rev) <- categories
  first <- T
  
  # Calculate incidence and reversal by category
  for (i in categories){
    # Obtain number of individuals stunted/not stunted
    stuntCount.inc <- pred.stunt %>% filter(visit==i & stunt==TRUE & cuuid %in% selectID.inc) %>% nrow
    stuntCount.rev <- pred.stunt %>% filter(visit==i & stunt==FALSE & cuuid %in% selectID.rev) %>% nrow
    
    # Ignore first column as we need two time points to calculate incidence/reversal
    if (first==T){
      out.inc[c(1,2),i] <- NA
      out.rev[c(1,2),i] <- NA
      first <- F
    }
    else{ # Obtain denominator and calculate incidence/reversal
      prevN.inc <- pred.stunt %>% filter(visit==i & cuuid %in% selectID.inc) %>% nrow()
      out.inc[1,i] <- stuntCount.inc/prevN.inc
      out.inc[2,i] <- prevN.inc
      
      prevN.rev <- pred.stunt %>% filter(visit==i & cuuid %in% selectID.rev) %>% nrow()
      out.rev[1,i] <- stuntCount.rev/prevN.rev
      out.rev[2,i] <- prevN.rev
    }
    
    # If baseline is null, only keep individuals in previous cohort 
    if (is.null(baseline)){
      selectID.inc <- pred.stunt %>% filter(visit==i & stunt==FALSE) %>% pull(cuuid)
      selectID.rev <- pred.stunt %>% filter(visit==i & stunt==TRUE) %>% pull(cuuid)
    }else{ # If baseline is specified, keep individuals from baseline
      selectID.inc <- pred.stunt %>% filter(visit==baseline & stunt==FALSE) %>% pull(cuuid)
      selectID.rev <- pred.stunt %>% filter(visit==baseline & stunt==TRUE) %>% pull(cuuid)
    }
  }
  return(cbind(tibble(Model=modelname,round(out.inc,3)), tibble(Model=modelname,round(out.rev,3))))
}




##### Generate tables #####
# Prevalence table
allPrev.2sd <- rbind(prevCalc(removesd(mod6,dat),"Mod6"),
                     prevCalc(dat,"Observed"))
allPrev.2sd %<>% pivot_wider(names_from = Category, values_from = Proportion)
allPrev.2sd # Get prevalence

# How many points from each model removed
removedPoints <- rbind(table(removesd(mod6,dat)$visit),
                       table(dat$visit))
removedPoints <- removedPoints %>% data.frame()
removedPoints$total <- apply(removedPoints, 1, sum)
colnames(removedPoints) <- c("Birth","Day 6","Day 28","Day 42","6 months","12 months","24 months","Total")
rownames(removedPoints) <- allPrev.2sd$Model
removedPoints # Get window counts

# Incidence and reversal tables
allIR.2sd <- rbind(inc_and_rev_Calc(removesd(mod6,dat),"Mod6"),
                   inc_and_rev_Calc(dat,"Observed"))

IR_n <- nrow(allIR.2sd)/2
inc <- allIR.2sd[rep(c(T,F),IR_n),1:5] # Get incidence
inc_c <- allIR.2sd[rep(c(F,T),IR_n),1:5] # Get incidence counts
rev <- allIR.2sd[rep(c(T,F),IR_n),6:10] # Get reversal
rev_c <- allIR.2sd[rep(c(F,T),IR_n),6:10] # Get reversal counts

# Incidence and reversal tables w/ common baseline of Day 6
allIR.2sd.baseline <- rbind(inc_and_rev_Calc(removesd(mod6,dat),"Mod6", "Day 6"),
                   inc_and_rev_Calc(dat,"Observed", "Day 6"))

IR_n.cb <- nrow(allIR.2sd.baseline)/2
inc.cb <- allIR.2sd.baseline[rep(c(T,F),IR_n.cb),1:5] # Get incidence w/ common baseline
inc_c.cb <- allIR.2sd.baseline[rep(c(F,T),IR_n.cb),1:5] # Get incidence counts w/ common baseline
rev.cb <- allIR.2sd.baseline[rep(c(T,F),IR_n.cb),6:10] # Get reversal w/ common baseline
rev_c.cb <- allIR.2sd.baseline[rep(c(F,T),IR_n.cb),6:10] # Get reversal counts w/ common baseline




##### Model diagnostics #####
AIC(mod1,mod2,mod3,mod4,mod5,mod6)
BIC(mod1,mod2,mod3,mod4,mod5,mod6)
sapply(c(mod1,mod2,mod3,mod4,mod5,mod6),deviance)




##### Proportion CI #####
# Returns Agresti-coull CI, given data frame with # of successes and # of trials
getCI <- function(data.in){
  colnames(data.in) <- c("Success","N")
  result <- round(binom.confint(data.in[,"Success"],data.in[,"N"],methods="agresti-coull")[4:6],3)*100
}

# Returns formatted dataset for plotting
createPlotCI <- function(n, N, categories){
  model <- cbind(type="Model",Age=categories, getCI(cbind(t(n[1,]), t(N[1,]))))
  obs <- cbind(type="Observed",Age=categories, getCI(cbind(t(n[2,]), t(N[2,]))))
  out <- rbind(model, obs)
  
  labels <-  c("Model", "Age", "Estimate", "Lower", "Upper")
  colnames(out) <- labels
  return(out)
}

# Confidence intervals for prevalence
categories <- factor(c("Birth","Day 6","Day 28","Day 42","6 months","12 months","24 months"), 
                     levels = c("Birth","Day 6","Day 28","Day 42","6 months","12 months","24 months"))
prev_dat <- createPlotCI(allPrev.2sd[,-1]*removedPoints[,-8], removedPoints[,-8], categories)

# Confidence intervals for sequential incidence 
inc_dat <- createPlotCI(inc[,3:5]*inc_c[,3:5], inc_c[,3:5], categories[5:7])

# Confidence intervals for common-baseline incidence 
inc_dat.cb <- createPlotCI(inc.cb[,3:5]*inc_c.cb[,3:5], inc_c.cb[,3:5], categories[5:7])

# Confidence intervals for sequential reversal 
rev_dat <- createPlotCI(rev[,3:5]*rev_c[,3:5], rev_c[,3:5], categories[5:7])

# Confidence intervals for common-baseline reversal 
rev_dat.cb <- createPlotCI(rev.cb[,3:5]*rev_c.cb[,3:5], rev_c.cb[,3:5], categories[5:7])




##### Plotting #####
# Scatter plot of age*length with age windows
al_scatter <- dat %>% mutate(age=age*30.437) %>% ggplot(aes(x=age, y=length)) + geom_point(shape=1, color="orange") +
  scale_x_continuous(name="Age (days)", breaks=c(0, 28, 180, 365, 730))
window_plot <- al_scatter +  theme(legend.position = "none") + ylab("Length") + geom_vline(xintercept = 165, linetype="dotted", color="Blue") +
  geom_vline(xintercept = 200, linetype="dotted", color="Blue") + geom_vline(xintercept = 350, linetype="dotted", color="Red") +
  geom_vline(xintercept = 380, linetype="dotted", color="Red") + geom_vline(xintercept = 715, linetype="dotted", color="green") +
  geom_vline(xintercept = 745, linetype="dotted", color="green") + geom_vline(xintercept = 40, linetype="dotted", color="yellow") +
  geom_vline(xintercept = 60, linetype="dotted", color="yellow") +geom_vline(xintercept = 20, linetype="dotted", color="black") +
  geom_vline(xintercept = 40, linetype="dotted", color="black") + geom_vline(xintercept = 12, linetype="dotted", color="cyan") +
  geom_vline(xintercept = 5, linetype="dotted", color="cyan") + geom_vline(xintercept = 0, linetype="dotted", color="pink") +
  geom_vline(xintercept = 4, linetype="dotted", color="pink") + ggtitle("Scatter plot of Age*Length with age windows ")

# # Scatter plot function of predicted vs. observed length
plotOP <- function(dat_in, modelText){
  dat_in %>% ggplot(aes(x=age, y=value, color=name)) + geom_point() +
    ggtitle(paste("Predicted vs. observed length",modelText)) +
    scale_color_manual(labels = c("Observed", "Predicted"), values = c("orange", "darkturquoise")) +
    labs(y="Length", x="Age",color="")
}

# Plot observed vs predicted length for model 6
# Create dataset with both observed and predicted values
pred.v.obs <- dat %>% select(age,age_s,length,visit,cuuid,n_obs) %>% mutate(p_length6=predict(mod6))
predicted_plot <- pred.v.obs %>% pivot_longer(c("length", "p_length6")) %>% plotOP("Model 6")

# Plot removed points for model 6
removed_pts <- anti_join(dat, removesd(mod6,dat)) %>% mutate(removed=T) %>% select(age,length,removed)
plot_pts <- removesd(mod6,dat) %>% mutate(removed=F) %>% select(age,length,removed) %>% rbind(removed_pts)
removed_plot <- ggplot(data=plot_pts, aes(x=age,y=length)) + geom_point(aes(color=removed, shape=removed)) + scale_shape_manual(values=c(16, 3)) + xlab("Age") + ylab("Length") + ggtitle("Points removed from model") 

# Plot proportions with confidence intervals
plotCI <- function(data.in, ylab, title){
  ggplot(data = data.in, aes(x = Age, y = Estimate, fill = Model, color = Model), size = 3) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        legend.title = element_blank(), legend.key = element_rect(fill = "white")) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.1,position=position_dodge(width=0.7)) +
  geom_point(position=position_dodge(width=0.7)) +
  geom_text(aes(label=format(round(Estimate,1),nsmall=1)),hjust=-0.2, size=3,show.legend = FALSE,position=position_dodge(width=0.7)) +
  scale_color_manual(values = c("#E7B800", "#2E9FDF")) + 
  scale_x_discrete(breaks = data.in$Age, labels = data.in$Age) +
  scale_y_continuous(limits = c(0,80), name = ylab) + ggtitle(title)
}

# Plot prevalence
prev_plot <- plotCI(prev_dat, ylab="Prevalence with 95% CI", title="Estimated prevalence of stunting")

# Plot sequential incidence
inc_plot <- plotCI(inc_dat, ylab="Incidence with 95% CI", title="Estimated incidence of stunting (Sequential)")

# Plot CB incidence
inc_plot.cb <- plotCI(inc_dat.cb, ylab="Incidence with 95% CI", title="Estimated incidence of stunting (Common baseline)")

# Plot sequential reversal
rev_plot <- plotCI(rev_dat, ylab="Reversal with 95% CI", title="Estimated incidence of reversal (Sequential)")

# Plot CB reversal
rev_plot.cb <- plotCI(rev_dat.cb, ylab="Reversal with 95% CI", title="Estimated incidence of reversal (Common baseline)")
```

# Data cleaning
For data cleaning two main steps were taken, leaving 10835 of the 19078 original observations:

1. First, invalid age and missing length observations were removed from the dataset. 
2. Second, facility visits after birth were removed from the dataset. 

Ages were categorized into 7 categories: 

* Birth (Days 0 to 3)
* Day 6 (Days 4 to 10)
* Day 28 (Days 20 to 35)
* Day 42 (Days 35 to 50)
* 6 months (Days 165 to 200)
* 12 months (Days 350 to 380)
* 24 months (Days 715 to 765)

# Modeling
Six mixed effects models were fit for the outcome of length. A combination of standardized age, quadratic age, and age splines were used as fixed and random effects.

```{r, eval=F}
# LME with random intercept
mod1 <- lmer(length ~ age_s + (1|cuuid), data=dat) 

# LME with random intercept+slope
mod2 <- lmer(length ~ age_s + (1+age_s|cuuid), data=dat) 

# LME with random intercept and quadratic age
mod3 <- lmer(length ~ age_s + I(age_s^2) + (1|cuuid), data=dat)

# LME with random intercept+slope and quadratic age
mod4 <- lmer(length ~ age_s + I(age_s^2) + (1 + age_s|cuuid), data=dat)

# LME with random intercept and piecewise-linear age
mod5 <- lmer(length ~ spline1 + spline2 + spline3 + spline4 + (1|cuuid), data=dat)

# LME with random intercept+slope and piecewise-linear age
mod6 <- lmer(length ~ spline1 + spline2 + spline3 + spline4 + (1 + age_s|cuuid), data=dat)
```

Out of these 6 models, we chose to use \textcolor{red}{model 6} which utilizes age splines for the fixed effects, and a random intercept and slope for age. We chose this model due to how the data was collected, as well as some model diagnostics (AIC, BIC, and deviance). Note that lower values indicate better fit:

```{r}
AIC(mod1,mod2,mod3,mod4,mod5,mod6)
BIC(mod1,mod2,mod3,mod4,mod5,mod6)
sapply(c(mod1,mod2,mod3,mod4,mod5,mod6),deviance)
```

# Measurement error
It was clear that measurement error was an issue in our dataset. To account for this, we attempted three correction techniques:

1. Removing observed lengths that were 1.5 standard deviations away from our predicted model lengths
2. Removing observations that did not increase monotonically (lengths should always increase over time)
3. For time windows with multiple observations, the average length was taken

Ultimately, we chose to utilize methods (1) and (3) for our calculations.

\newpage
# Visualizing windows and removed points
To visualize the windows of our age categories:

```{r}
window_plot
```

\newpage
In order to determine which points to remove from our dataset, we compared the model's predicted points with the observed dataset:

```{r}
predicted_plot
```

\newpage
We removed observed points that were 1.5 standard deviations away from our predicted points:

```{r}
removed_plot
```

When using our model to remove points from our dataset, we ended up with the following counts in each window:

```{r}
removedPoints
```

\newpage
# Prevalence, incidence, and reversal
After removing the relevant points in our dataset, we used the remaining data points to calculate our estimates of prevalence, incidence, and reversal of stunting.

## Prevalence:

```{r}
prev_dat # Get prevalence
```


## Incidence and reversal:
Incidence and reversal was calculated using only 4 categories: Day 6, 6 months, 12 months, and 24 months.
Furthermore, incidence and reversal was calculated in two ways.

1. Sequential - (Day 6 vs 6 months, 6 months vs 12 months, 12 months vs 24 months)
2. Common baseline - Day 6 (Day 6 vs 6 months, Day 6 vs 12 months, Day 6 vs 24 months)

```{r}
inc_dat # Get sequential incidence
rev_dat # Get sequential reversal
inc_dat.cb # Get incidence w/ common baseline
rev_dat.cb # Get reversal w/ common baseline
```

## Incidence and reversal counts:
Depending on the baseline value, the number of people in the "risk set" changes:

```{r}
inc_c # Get sequential incidence counts 
rev_c # Get sequential reversal counts
inc_c.cb # Get incidence counts w/ common baseline
rev_c.cb # Get reversal counts w/ common baseline
```

# Plotting incidence and reversal

```{r, echo=FALSE, fig.width=9}
prev_plot # Prevalence plot
inc_plot # Sequential incidence plot
rev_plot # Sequential reversal plot 
inc_plot.cb # Incidence plot with common baseline
rev_plot.cb # Reversal plot with common baseline
```








